::: {.callout-note appearance="simple"}
### **Jesus C. Compaire** \| {{< fa fish >}} {{< fa fish >}} {{< fa fish >}} | [{{< fa envelope >}}](mailto:jesus.canocompaire@uca.es) \| [{{< fa brands github >}}](https://github.com/jcompaire)
Assistant Researcher - **CESIMAR (CONICET)** \| Assistant Professor - **UNPSJB**
:::

### **Description**

Code to replicate the statistical analyses and generate figures performed in the section **3.3 Generalzed Linear and Additive Models** of the manuscript:

Compaire, J.C., Irigoyen, A.J. Simionato C.G. & Acha, E.M. (2026). Multiscale environmental forcing of multispecies fishery dynamics across climate timescales. Submitted to *Fisheries Oceanography*, **XXX** <https://doi.org/XXXXX> \|

---
title: " Code 2 of 2 - Multiscale fishery variability"
date: "23-04-2026"
---

# **Set console, environment, packages, directory, and functions**

```{r, results='hide'}
##| results: hide
# Clear console, environment, figures and free memory
cat("\014"); rm(list=ls()); graphics.off(); gc()
# Load packages needed for analysis and figures
pk <- c(
  "comprehenr", "crayon", "dtwclust", "dplyr", "forecast", "kableExtra",
  "knitr", "ggplot2", "ggpubr", "ggspatial", "ggthemes", "gratia", "grid",
  "gridExtra", "gridtext", "MARSS", "maptools", "metR", "mgcv", "ncdf4",
  "patchwork", "performance", "plyr", "purrr", "raster", "readxl", "reshape2",
  "rnaturalearth", "sf", "scales", "stats", "stringr", "tidyr", "tidyverse"
  )
invisible(lapply(pk, require, character.only = TRUE))
# Set working directory, and invoke functions
source("dtw_dfa_glm_gam_aucfz_functions.R") # functions
out_dir_fig <- "outputs/figures"
out_dir_files <- "outputs/files"
if (!dir.exists(out_dir_fig)) dir.create(out_dir_fig, recursive = TRUE)
if (!dir.exists(out_dir_files)) dir.create(out_dir_files, recursive = TRUE)
```

# **Read environmental data**

```{r, results='hold'}
##| results: hold
nc <- nc_open(paste0("datasets/Environmental_lagged_arrays_10years.nc")) # open ncfile
nc_info <- capture.output(print(nc))
cat(nc_info[-1], sep = "\n")
env_out <- read_env_vars(nc)
env_var_lag <- env_out$env_var_lag
time        <- env_out$time
lag         <- c(1:length(env_var_lag)-1)
nc_close(nc) # close ncfile
```

# **GLMs analyses for i) DTW-centroids and ii) DFA-states**

## Load DTW-centroids data

```{r}
# --- Read centroids
load(paste0(out_dir_files, '/centroids_DBA.RData')) # centroids
fish_df <- get_states_or_centroids(t(dba_centroids), years = 1996:2023,
                                   tag = 'centroids')
# --- Combine names of environmental and fishing datasets
names_col <- c(env_out$names_col, colnames(fish_df)[-length(fish_df)])
# --- Merge environmental and fishing data by lag
env_fish_lag_dtw <- merge_env_states(env_var_lag, fish_df, names_col)
```

## Getting all possible formulas

```{r, results='asis'}
#  --- Generate all possible formulas
th <- c('25','50','75') # percentiles
response <- "y"  # generic name for response variable
glms_formulas_by_ix <- list()
for(k in 1:length(th)){
  # List of variables and index selection
  vq <- c(th[k], response)
  idx_var <- to_vec(
    for(i in 1:length(vq)) str_which(names_col, vq[i], negate = FALSE)
  )
  names_var <- c('soi','sam','enso', names_col[idx_var])
  
  #  Combinations of climate indices
  ci <- c('soi','sam','enso')
  cixs <- to_list(
    for(i in 1:length(ci)) c(combn(ci, i, simplify = FALSE))
  )
  
  cixss <- list()
  cixss[[1]] <- cixs[[2]]   # pairs
  cixss[[1]][[4]] <- unlist(cixs[[3]]) # all
  
  for(ix in 1:length(cixss[[1]])){
    cith <- paste(paste(cixss[[1]][[ix]], collapse = '|'), sep = '|')
    names_var_ix <- names_var[-grep(pattern = cith, names_var)]
    
    #  Remove percentile tag
    for(tag in th[k]) names_var_ix <- gsub(tag,'', names_var_ix)
    
    # Covariates and response
    xnames <- names_var_ix
    ynames <- response
    
    # Formulas with interactions
    inter <- c("riv", "kd", "sss")
    glms_f <- comb_glmi(xnames, ynames, inter)
    
    if(ix < length(cixss[[1]])){
      glms_f <- glms_f[grep(pattern = paste(cixss[[1]][[4]], collapse = '|'), glms_f)]
    }
    glms_formulas_by_ix[[ix]] <- glms_f
  }
}
# Combine all formulas into a single object
glms_formulas <- do.call(c, glms_formulas_by_ix)
# Transform to character
# glms_formulas_chr <- sapply(
#   glms_formulas, function(f) paste(deparse(f), collapse = " ")
#   )
cat(sapply(head(glms_formulas, 3), format),
    "... and so on until... ",
    sapply(tail(glms_formulas, 3), format), sep = "\n\n")
```

## Run GLMs for DTW-centroids

```{r}
glm_dtw_results <- run_glms_nested(
  nvariables = 4,
  env_fish_lag = env_fish_lag_dtw,
  lag = lag, # lags
  th = th, # percentiles
  glms_formulas = glms_formulas, # list
  fmly = list(quasi(link = "identity", variance = "constant")), # family
  add_th_suffix = add_th_suffix, # function
  try_glm_stepwise = try_glm_stepwise # function
)
```

## Save summaries of significant models

```{r}
outfile <- paste0(out_dir_files, '/SupplementaryMaterial_S12_GLMs_DTW.txt')
sink(outfile, append = FALSE)
print_glm_nested_results(glm_dtw_results)
sink()
closeAllConnections()
```

## Load DFA states data

```{r}
# --- Read states
load(paste0(out_dir_files, '/states_DFA.RData')) # states
fish_df <- get_states_or_centroids(states_dfa, years = 1996:2023,
                                   tag = 'states')
# --- Combine names of environmental and fishing datasets
names_col <- c(env_out$names_col, colnames(fish_df)[-length(fish_df)])
# --- Merge environmental and fishing data by lag
env_fish_lag_dfa <- merge_env_states(env_var_lag, fish_df, names_col)
```

## Run GLMs for DFA-states

```{r}
glm_dfa_results <- run_glms_nested(
  nvariables = 3,
  env_fish_lag = env_fish_lag_dfa,
  lag = lag, # lags
  th = th, # percentiles
  glms_formulas = glms_formulas, # list
  fmly = list(quasi(link = "identity", variance = "constant")), # family
  add_th_suffix = add_th_suffix, # function
  try_glm_stepwise = try_glm_stepwise # function
  )
```

## Save summaries of significant models

```{r}
outfile <- paste0(out_dir_files, '/SupplementaryMaterial_S13_GLMs_DFA.txt')
sink(outfile, append = FALSE)
print_glm_nested_results(glm_dfa_results)
sink()
closeAllConnections()
```

## Save the outputs as R dataset objects

```{r}
# For DTW centroids as response variable
save(
  glm_dtw_results,
  file = file.path(out_dir_files,'/GLMs_DTW_results.RData'))
# For DFA states as response variable
save(
  glm_dfa_results,
  file = file.path(out_dir_files, "/GLMs_DFA_results.RData"))
```

# **Load GLMs results, create subplot elements and plot lags and DE for centroids and common hidden trends**

```{r, results = 'asis', fig.width=12, fig.height=9}
# --- Loading results for DFA states and DTW centroids
load(paste0(out_dir_files, '/GLMs_DTW_results.RData'))
load(paste0(out_dir_files, '/GLMs_DFA_results.RData'))
# Centroids
c1 <- tables_states_glm(glm_dtw_results[['c1']], '1')
c2 <- tables_states_glm(glm_dtw_results[['c2']], '2')
c3 <- tables_states_glm(glm_dtw_results[['c3']], '3')
c4 <- tables_states_glm(glm_dtw_results[['c4']], '4')
# States
s1 <- tables_states_glm(glm_dfa_results[['s1']], '1')
s2 <- tables_states_glm(glm_dfa_results[['s2']], '2')
s3 <- tables_states_glm(glm_dfa_results[['s3']], '3')

# --- Getting list by percentile
percentiles <- c("th25", "th50", "th75")
# --- Function calls "states" and "centroids" lists (s1, s2, s3, and c1, c2, c3, c4)
cent_percentiles_list <- combine_by_percentiles(percentiles, prefix = "c", n_elements = 4)
stat_percentiles_list <- combine_by_percentiles(percentiles, prefix = "s", n_elements = 3)

# --- Getting each percentile as dataframe for "states" and "centroids"
# Centroids
cent_th25 <- cent_percentiles_list[["th25"]]
cent_th50 <- cent_percentiles_list[["th50"]]
cent_th75 <- cent_percentiles_list[["th75"]]
# States
stat_th25 <- stat_percentiles_list[["th25"]]
stat_th50 <- stat_percentiles_list[["th50"]]
stat_th75 <- stat_percentiles_list[["th75"]]

# --- Create each subplot element
# Centroids
c25 <- plot_de_lag_annotated(cent_th25, "25th", "A", fill_label = NULL)
c50 <- plot_de_lag_annotated(cent_th50, "50th", "B", fill_label = NULL)
c75 <- plot_de_lag_annotated(cent_th75, "75th", "C", show_legend = TRUE, fill_label = "Centroids", 0.8, 0.87)
# States
s25 <- plot_de_lag_annotated(stat_th25, "25th", "A", fill_label = NULL)
s50 <- plot_de_lag_annotated(stat_th50, "50th", "B", fill_label = NULL)
s75 <- plot_de_lag_annotated(stat_th75, "75th", "C", show_legend = TRUE, fill_label = "Trends", 0.92, 0.9)

# Merging plots into a single one and export it as PDF
# Centroids
file_path <- file.path(out_dir_fig, "Fig_S8")
# --- as PDF
pdf(paste0(file_path, ".pdf"), width = 12, height = 9, colormodel = "cmyk")
par(mfcol=c(1,1))
figure <- ggarrange(c25, c50, c75,
                    ncol = 3, nrow = 1, common.legend = F)
f_dtw <- annotate_figure(
  figure, 
  bottom = textGrob(
    expression(paste("Lag (years)")),
    rot = 0, vjust = 0.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  ),
  left = textGrob(
    expression(paste("Deviance explained (%)")),
    rot = 90, vjust = 1.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  )) 
print(f_dtw)
invisible(dev.off())
# --- as PNG
ragg::agg_png(paste0(file_path, ".png"), width = 12, height = 9, units = "in", res = 600)
print(f_dtw)
invisible(dev.off())
# HTML output
if (knitr::is_html_output()) {
  cat("\n\n")
  print(f_dtw)
  cat("\n\n**Figure S8.-** Average (triangles) and maximum (bars) deviance explained (%) by GLMs for the DTW-derived centroids 1-4 at different time lags (from 0 to 10 years). Numbers above the bars indicate the count of significant models for each specific lag. \n\n")
}

# States
file_path <- file.path(out_dir_fig, "Fig_S9")
# --- as PDF
pdf(paste0(file_path, ".pdf"), width = 12, height = 9, colormodel = "cmyk")
par(mfcol=c(1,1))
figure <- ggarrange(s25, s50, s75,
                    ncol = 3, nrow = 1, common.legend = F)
f_dtw <- annotate_figure(
  figure, 
  bottom = textGrob(
    expression(paste("Lag (years)")),
    rot = 0, vjust = 0.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  ),
  left = textGrob(
    expression(paste("Deviance explained (%)")),
    rot = 90, vjust = 1.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  )) 
print(f_dtw)
invisible(dev.off())
# --- as PNG
ragg::agg_png(paste0(file_path, ".png"), width = 12, height = 9, units = "in", res = 600)
print(f_dtw)
invisible(dev.off())
# HTML output
if (knitr::is_html_output()) {
  cat("\n\n")
  print(f_dtw)
  cat("\n\n**Figure S9.-** Average (triangles) and maximum (bars) deviance explained (%) by GLMs for the DFA-derived common hidden trends at different time lags (from 0 to 10 years). Numbers above the bars indicate the count of significant models for each specific lag. \n\n\n\n
  To get the estimated parameters of the explanatory variables presented in Table 2 it is necessary to check the results saved in 'glm_dtw_results' and 'glm_dfa_results'.")
}

```

## GLMs with highest Deviance Explained for DTW-centroids and DFA-states

```{r, echo=FALSE}
# Checking model residuals for DTW with highest DE, and plotting
# --- Centroid 1 
cat("===============  Centroid 1 ===============")
model_dtw <- glm("c1 ~ riv50 + kd50 + sss50 + riv50:kd50 + riv50:sss50",
                 family = quasi(link="identity", variance="constant"),
                 data = env_fish_lag_dtw[[4]], na.action = na.omit)

summary(model_dtw)
de_dtw <- round((1 - model_dtw[["deviance"]] / model_dtw[["null.deviance"]]) * 100, digits = 2)
check_collinearity(model_dtw)
shapiro.test(model_dtw$residuals)
var_contrib_from_full(model_dtw)
# --- Centroid 2
cat("\n\n =============== Centroid 2 ===============")
model_dtw <- glm("c2 ~ kd75 + sss75",
                 family = quasi(link="identity", variance="constant"),
                 data = env_fish_lag_dtw[[2]], na.action = na.omit)

summary(model_dtw)
de_dtw <- round((1 - model_dtw[["deviance"]] / model_dtw[["null.deviance"]]) * 100, digits = 2)
check_collinearity(model_dtw)
shapiro.test(model_dtw$residuals)
var_contrib_from_full(model_dtw)
# Figure 
effects_dtw_glm <- compute_glm_all_effects(model_dtw)
vrs <- names(effects_dtw_glm) # Should be c("kd75", "sss75")
col_line <- c("#E69F00", "darkgreen") # Custom colors
plot_list <- list() # We create the list to store the plots
tag_positions <- list(c(0.35, 0.975), c(0.21, 0.975))
for(i in seq_along(vrs)) { # iterate from 1 to 3 (or the number of variables)
  # Get data for this iteration
  var_selected <- vrs[i]
  color_selected <- col_line[i]
  # Get the prediction data (from the 'effects_glm' list)
  data_pred <- effects_dtw_glm[[var_selected]] 
  # Get the original data (for the rug plot)
  data_model <- model_dtw$data            
  # Call the plotting function
  p <- plot_glm_effect(
    prediction_data = data_pred,
    model_data = data_model,
    focal_var = var_selected,
    line_color = color_selected,
    y_label = if (i == 1) "Predicted Value\n(DBA Centroid 2)" else NULL

    )
  # Add the tag (LETTERS)
  plot_list[[var_selected]] <- p + labs(tag = LETTERS[i+2]) +
    theme(
      plot.tag.position = tag_positions[[i]],
      plot.tag = element_text(
        face = "bold", size = 20)
    )
}
# Saving figure elements
plot_GLM_c2 <- plot_list
save(plot_GLM_c2,
  file = paste0(out_dir_files, '/Fig_8_CD_c2_GLM.RData')
  )
# --- Centroid 3
cat("\n\n =============== Centroid 3 ===============")
model_dtw <- glm("c3 ~ riv25 + sss25 + riv25:sss25",
                 family = quasi(link="identity", variance="constant"),
                 data = env_fish_lag_dtw[[4]], na.action = na.omit)

summary(model_dtw)
de_dtw <- round((1 - model_dtw[["deviance"]] / model_dtw[["null.deviance"]]) * 100, digits = 2)
check_collinearity(model_dtw)
shapiro.test(model_dtw$residuals)
var_contrib_from_full(model_dtw)
# Figure
# --- GLM single variables 
effects_dtw_glm <- compute_glm_all_effects(model_dtw)
vrs <- names(effects_dtw_glm) # Should be c("riv25", "sss75")
col_line <- c("#56B4E9", "darkgreen") # Custom colors
plot_list <- list() # We create the list to store the plots
tag_positions <- list(c(0.35, 0.975), c(0.21, 0.975), c(0.21, 0.975))
for(i in seq_along(vrs)) { # iterate from 1 to 3 (or the number of variables)
  # Get data for this iteration
  var_selected <- vrs[i]
  color_selected <- col_line[i]
  # Get the prediction data (from the 'effects_glm' list)
  data_pred <- effects_dtw_glm[[var_selected]] 
  # Get the original data (for the rug plot)
  data_model <- model_dtw$data            
  # Call the plotting function
  p <- plot_glm_effect(
    prediction_data = data_pred,
    model_data = data_model,
    focal_var = var_selected,
    line_color = color_selected,
    y_label = if (i == 1) "Predicted Value\n(DBA Centroid 3)" else NULL
    )
  # Add the tag (LETTERS)
  plot_list[[var_selected]] <- p + labs(tag = LETTERS[i+4]) +
    theme(
      plot.tag.position = tag_positions[[i]],
      plot.tag = element_text(
        face = "bold", size = 20)
    )
}
# --- GLM interaction variables 
pi <- plot_glm_interaction(
  model = model_dtw,
  model_data = env_fish_lag_dtw[[4]],
  focal_var = "riv25",
  moderator_var = "sss25",
  probs = c(0.1, 0.9),   # custom percentiles used inside predict_terms_auto()
  custom_colors = c(
    adjustcolor("darkseagreen", alpha.f = 1),
    adjustcolor("darkgreen", alpha.f = 1)
    ),
  y_label = "Predicted Value\n(DBA Centroid 3)"
  )
plot_list[["riv25:sss25"]] <- pi + labs(tag = LETTERS[i+4+3]) +
    theme(
      plot.tag.position = c(0.17, 0.975),
      plot.tag = element_text(
        face = "bold", size = 20)
    )
pi2 <- plot_glm_interaction(
  model = model_dtw,
  model_data = env_fish_lag_dtw[[4]],
  focal_var = "sss25",
  moderator_var = "riv25",
  probs = c(0.1, 0.95),   # custom percentiles used inside predict_terms_auto()
  custom_colors = c(
    adjustcolor("lightblue", alpha.f = 1),
    adjustcolor("dodgerblue", alpha.f = 1)
  ),
  y_label = "Predicted Value\n(DBA Centroid 3)"
)
plot_list[["sss25:riv25"]] <- pi2 + ylab(NULL) + labs(tag = LETTERS[i+5]) +
    theme(
      plot.tag.position = tag_positions[[3]],,
      plot.tag = element_text(
        face = "bold", size = 20)
    )
# Saving figure elements
plot_GLM_c3 <- plot_list
save(plot_GLM_c3,
  file = paste0(out_dir_files, '/Fig_8_EFG_c3_GLM.RData')
  )
# --- Centroid 4
cat("\n\n =============== Centroid 4 ===============")
model_dtw <- glm("c4 ~ soi + riv75 + kd75",
                 family = quasi(link="identity", variance="constant"),
                 data = env_fish_lag_dtw[[11]], na.action = na.omit)

summary(model_dtw)
de_dtw <- round((1 - model_dtw[["deviance"]] / model_dtw[["null.deviance"]]) * 100, digits = 2)
check_collinearity(model_dtw)
shapiro.test(model_dtw$residuals)
var_contrib_from_full(model_dtw)

# Checking model residuals for DFA with highest DE, and plotting
# --- State 1
cat("\n\n =============== Common hidden trend 1 ===============")
model_dfa <- glm("s1 ~ riv75 + kd75 + wmi75",
                 family = quasi(link="identity", variance="constant"),
                 data = env_fish_lag_dfa[[7]], na.action = na.omit)

summary(model_dfa)
round((1 - model_dfa[["deviance"]] / model_dfa[["null.deviance"]]) * 100, digits = 2)
check_collinearity(model_dfa)
shapiro.test(model_dfa$residuals)
(var_contrib_from_full(model_dfa))
# --- State 2
cat("\n\n =============== Common hidden trend 2 ===============")
model_dfa <- glm("s2 ~ sam + sst25 + kd25",
                 family = quasi(link="identity", variance="constant"),
                 data = env_fish_lag_dfa[[8]], na.action = na.omit)

summary(model_dfa)
round((1 - model_dfa[["deviance"]] / model_dfa[["null.deviance"]]) * 100, digits = 2)
check_collinearity(model_dfa)
shapiro.test(model_dfa$residuals)
(var_contrib_from_full(model_dfa))
# Figure
effects_dfa_glm <- compute_glm_all_effects(model_dfa)
vrs <- names(effects_dfa_glm)
col_line <- c("#FFB6C1", "#FF6347", "#E69F00") # sam, ssst, kd
plot_list <- list() # We create the list to store the plots
tag_positions <- list(c(0.35, 0.975), c(0.21, 0.975), c(0.21, 0.975))
for(i in seq_along(vrs)) { # iterate from 1 to 3 (or the number of variables)
  # Get data for this iteration
  var_selected <- vrs[i]
  color_selected <- col_line[i]
  # Get the prediction data (from the 'effects_glm' list)
  data_pred <- effects_dfa_glm[[var_selected]] 
  # Get the original data (for the rug plot)
  data_model <- model_dfa$data            
  # Call the plotting function
  p <- plot_glm_effect(
    prediction_data = data_pred,
    model_data = data_model,
    focal_var = var_selected,
    line_color = color_selected,
    y_label = if (i == 1) "Predicted Value\n(DFA Common trend 2)" else NULL 
  )
  # Add the tag (LETTERS)
  plot_list[[var_selected]] <- p + labs(tag = LETTERS[i+2]) +
    theme(
      plot.tag.position = tag_positions[[i]],
      plot.tag = element_text(
        face = "bold", size = 20)
    )
}
# Saving figure elements
plot_GLM_s2 <- plot_list
save(plot_GLM_s2,
  file = paste0(out_dir_files, '/Fig_9_CDE_s2_GLM.RData')
  )

# --- State 3
cat("\n\n =============== Common hidden trend 3 ===============")
model_dfa <- glm(s3 ~ soi + riv75 + kd75 + wmi75,
                 family = quasi(link="identity", variance="constant"),
                 data = env_fish_lag_dfa[[7]], na.action = na.omit)

summary(model_dfa)
round((1 - model_dfa[["deviance"]] / model_dfa[["null.deviance"]]) * 100, digits = 2)
check_collinearity(model_dfa)
shapiro.test(model_dfa$residuals)
(var_contrib_from_full(model_dfa))
# Figure 
effects_dfa_glm <- compute_glm_all_effects(model_dfa)
vrs <- names(effects_dfa_glm)
col_line <- c("#492050", "#56B4E9", "#E69F00", "#000000") # Custom colors
plot_list <- list() # We create the list to store the plots
tag_positions <- list(
  c(0.35, 0.975), c(0.21, 0.975), c(0.21, 0.975), c(0.21, 0.975)
  )
for(i in seq_along(vrs)) { # iterate from 1 to 4 (or the number of variables)
  # Get data for this iteration
  var_selected <- vrs[i]
  color_selected <- col_line[i]
  # Get the prediction data (from the 'effects_glm' list)
  data_pred <- effects_dfa_glm[[var_selected]] 
  # Get the original data (for the rug plot)
  data_model <- model_dfa$data            
  # Call the plotting function
  p <- plot_glm_effect(
    prediction_data = data_pred,
    model_data = data_model,
    focal_var = var_selected,
    line_color = color_selected,
    y_label = if (i == 1) "Predicted Value\n(DFA Common trend 3)" else NULL 
  )
  # Add the tag (LETTERS)
  plot_list[[var_selected]] <- p + labs(tag = LETTERS[i+5]) +
    theme(
      plot.tag.position = tag_positions[[i]],
      plot.tag = element_text(
        face = "bold", size = 20)
    )
}
# Saving figure elements
plot_GLM_s3 <- plot_list
save(plot_GLM_s3,
  file = paste0(out_dir_files, '/Fig_9_FGHI_s3_GLM.RData')
  )
```

# **GAMs analyses for i) DTW-centroids and ii) DFA-states**

## Getting all possible formulas

```{r, results='asis'}
##| results: asis
th <- c('25','50','75') # percentiles
response <- "y"  # generic name for response variable
gams_formulas_by_ix <- list()
for(k in 1:length(th)){
  # List of variables and index selection
  vq <- c(th[k], response)
  idx_var <- to_vec(
    for(i in 1:length(vq)) str_which(names_col, vq[i], negate = FALSE)
  )
  names_var <- c('soi','sam','enso', names_col[idx_var])
  
  #  Combinations of climate indices
  ci <- c('soi','sam','enso')
  cixs <- to_list(
    for(i in 1:length(ci)) c(combn(ci, i, simplify = FALSE))
  )
  
  cixss <- list()
  cixss[[1]] <- cixs[[2]]   # pairs
  cixss[[1]][[4]] <- unlist(cixs[[3]]) # all
  
  for(ix in 1:length(cixss[[1]])){
    cith <- paste(paste(cixss[[1]][[ix]], collapse = '|'), sep = '|')
    names_var_ix <- names_var[-grep(pattern = cith, names_var)]
    
    #  Remove percentile tag
    for(tag in th[k]) names_var_ix <- gsub(tag,'', names_var_ix)
    
    # Covariates and response
    xnames <- names_var_ix
    ynames <- response
    
    # Formulas with interactions
    inter <- c("riv", "kd", "sss")
    gams_f <- comb_gami(xnames, ynames, 4, inter)
    
    if(ix < length(cixss[[1]])){
      gams_f <- gams_f[grep(pattern = paste(cixss[[1]][[4]], collapse = '|'), gams_f)]
    }
    gams_formulas_by_ix[[ix]] <- gams_f
  }
}
# Combine all formulas into a single object
gams_formulas <- do.call(c, gams_formulas_by_ix)
# Transform to character
# gams_formulas_chr <- sapply(
#   gams_formulas, function(f) paste(deparse(f), collapse = " ")
#   )
clean_format <- function(x) paste(format(x), collapse = "")
head_f <- sapply(head(gams_formulas, 3), clean_format)
tail_f <- sapply(tail(gams_formulas, 3), clean_format)
cat(paste(head_f, collapse = "\n\n"),
    "\n\n... and so on until... \n\n",
    paste(tail_f, collapse = "\n\n"),
    sep = ""
    )
```

## Run GAMs for DTW-centroids

```{r}
set.seed(23)
suppressWarnings({
gam_dtw_results <- run_gams_nested(
  nvariables = 4,
  env_fish_lag = env_fish_lag_dtw,
  lag = lag, # lags
  th = th, # percentiles
  gams_formulas = gams_formulas, # list
  fmly = list(quasi(link = "identity", variance = "constant")), # family
  add_th_suffix = add_th_suffix, # function
  try_gam_stepwise = try_gam_stepwise, # function
  p_threshold = 0.05, # ADDED
  concurvity_threshold = 0.3, # ADDED
  method = "REML")
})
```

## Save summaries of significant models

```{r}
outfile <- paste0(out_dir_files, '/SupplementaryMaterial_S14_GAMs_DTW.txt')
sink(outfile, append = FALSE)
print_gam_nested_results(gam_dtw_results)
sink()
closeAllConnections()
```

## Run GAMs for DFA-states

```{r}
set.seed(23)
suppressWarnings({
gam_dfa_results <- run_gams_nested(
  nvariables = 3,
  env_fish_lag = env_fish_lag_dfa,
  lag = lag, # lags
  th = th, # percentiles
  gams_formulas = gams_formulas, # list
  fmly = list(quasi(link = "identity", variance = "constant")), # family
  add_th_suffix = add_th_suffix, # function
  try_gam_stepwise = try_gam_stepwise, # function
  p_threshold = 0.05, # ADDED
  concurvity_threshold = 0.3, # ADDED
  method = "REML")
})
```

## Save summaries of significant models

```{r}
outfile <- paste0(out_dir_files, '/SupplementaryMaterial_S15_GAMs_DFA.txt')
sink(outfile, append = FALSE)
print_gam_nested_results(gam_dfa_results)
sink()
closeAllConnections()
```

## Save the outputs as R dataset objects

```{r}
# For DTW centroids as response variable
save(
  gam_dtw_results,
  file = file.path(out_dir_files,'GAMs_DTW_results.RData')
  )
# For DFA states as response variable
save(
  gam_dfa_results,
  file = file.path(out_dir_files, 'GAMs_DFA_results.RData')
  )
```

# **Load GAMs results, create subplot elements and plot lags and DE for centroids and common hidden trends**

```{r, results = 'asis', fig.width=12, fig.height=9}
# --- Loading results for DFA states and DTW centroids
load(paste0(out_dir_files, '/GAMs_DTW_results.RData'))
load(paste0(out_dir_files, '/GAMs_DFA_results.RData'))
# Centroids
c1 <- tables_states_gam(gam_dtw_results[['c1']], '1')
c2 <- tables_states_gam(gam_dtw_results[['c2']], '2')
c3 <- tables_states_gam(gam_dtw_results[['c3']], '3')
c4 <- tables_states_gam(gam_dtw_results[['c4']], '4')
# States
s1 <- tables_states_gam(gam_dfa_results[['s1']], '1')
s2 <- tables_states_gam(gam_dfa_results[['s2']], '2')
s3 <- tables_states_gam(gam_dfa_results[['s3']], '3')
# --- Getting list by percentile
percentiles <- c("th25", "th50", "th75")
# --- Function calls "states" and "centroids" lists (s1, s2, s3, and c1, c2, c3, c4)
stat_percentiles_list <- combine_by_percentiles(percentiles, prefix = "s", n_elements = 3)
cent_percentiles_list <- combine_by_percentiles(percentiles, prefix = "c", n_elements = 4)

# --- Getting each percentile as dataframe for "states" and "centroids"
# Centroids
cent_th25 <- cent_percentiles_list[["th25"]]
cent_th50 <- cent_percentiles_list[["th50"]]
cent_th75 <- cent_percentiles_list[["th75"]]
# States
stat_th25 <- stat_percentiles_list[["th25"]]
stat_th50 <- stat_percentiles_list[["th50"]]
stat_th75 <- stat_percentiles_list[["th75"]]

# --- Create each subplot element
# Centroids
c25 <- plot_de_lag_annotated(cent_th25, "25th", "A", fill_label = NULL)
c50 <- plot_de_lag_annotated(cent_th50, "50th", "B", fill_label = NULL)
c75 <- plot_de_lag_annotated(cent_th75, "75th", "C", show_legend = TRUE, fill_label = "Centroids", 0.7, 0.87)
# States
s25 <- plot_de_lag_annotated(stat_th25, "25th", "A", fill_label = NULL)
s50 <- plot_de_lag_annotated(stat_th50, "50th", "B", fill_label = NULL)
s75 <- plot_de_lag_annotated(stat_th75, "75th", "C", show_legend = TRUE, fill_label = "Trends", 0.92, 0.9)

# Merging plots into a single one and export it as PDF
# Centroids
file_path <- file.path(out_dir_fig, "Fig_S10")
# --- as PDF
pdf(paste0(file_path, ".pdf"), width = 12, height = 9, colormodel = "cmyk")
par(mfcol=c(1,1))
figure <- ggarrange(c25, c50, c75,
                    ncol = 3, nrow = 1, common.legend = F)
f_dtw <- annotate_figure(
  figure, 
  bottom = textGrob(
    expression(paste("Lag (years)")),
    rot = 0, vjust = 0.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  ),
  left = textGrob(
    expression(paste("Deviance explained (%)")),
    rot = 90, vjust = 1.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  )) 
print(f_dtw)
invisible(dev.off())
# --- as PNG
ragg::agg_png(paste0(file_path, ".png"), width = 12, height = 9, units = "in", res = 600)
print(f_dtw)
invisible(dev.off())
# HTML output
if (knitr::is_html_output()) {
  cat("\n\n")
  print(f_dtw)
  cat("\n\n**Figure S10.-** Average (triangles) and maximum (bars) deviance explained (%) by GAMs for the DTW-derived centroids 1-4 at different time lags (from 0 to 10 years). Numbers above the bars indicate the count of significant models for each specific lag. \n\n")
}

# States
file_path <- file.path(out_dir_fig, "Fig_S11")
# --- as PDF
pdf(paste0(file_path, ".pdf"), width = 12, height = 9, colormodel = "cmyk")
par(mfcol=c(1,1))
figure <- ggarrange(s25, s50, s75,
                    ncol = 3, nrow = 1, common.legend = F)
f_dtw <- annotate_figure(
  figure, 
  bottom = textGrob(
    expression(paste("Lag (years)")),
    rot = 0, vjust = 0.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  ),
  left = textGrob(
    expression(paste("Deviance explained (%)")),
    rot = 90, vjust = 1.3,
    gp = gpar(cex = 1.5, fontfamily = "Times" , fontsize = 12)
  )) 
print(f_dtw)
invisible(dev.off())
# --- as PNG
ragg::agg_png(paste0(file_path, ".png"), width = 12, height = 9, units = "in", res = 600)
print(f_dtw)
invisible(dev.off())
# HTML output
if (knitr::is_html_output()) {
  cat("\n\n")
  print(f_dtw)
  cat("\n\n**Figure S11.-** Average (triangles) and maximum (bars) deviance explained (%) by GAMs for the DFA-derived common hidden trends at different time lags (from 0 to 10 years). Numbers above the bars indicate the count of significant models for each specific lag.\n\n\n\n
  To get the estimated parameters of the explanatory variables presented in Table 3 it is necessary to check the results saved in 'gam_dtw_results' and 'gam_dfa_results.'")
}
```

## GAMs with highest Deviance Explained for DTW-centroids and DFA-states

```{r, results='hold'}
set.seed(123)
# Checking model residuals for DTW with highest DE, and plotting
# --- Centroid 1
cat("=============== Centroid 1 ===============")
model_dtw <- gam(
  (c1 ~ s(sst25, k = 4, bs = "cr") + s(kd25, k = 4, bs = "cr")),
  family = quasi(link="identity", variance="constant"),
  data = env_fish_lag_dtw[[4]], na.action = na.omit, method = "REML"
  )
summary(model_dtw)
de_dtw <- round((1 - model_dtw[["deviance"]] / model_dtw[["null.deviance"]]) * 100, digits = 2)
concurvity(model_dtw)
shapiro.test(model_dtw$residuals)
var_contrib_from_full_gam(model_dtw)
# Figure 
vrs <- names(model_dtw[["var.summary"]]) # 
col_line <- c("#FF6347", "#E69F00") # sst, kd
plot_list <- list() # We create the list to store the plots
tag_positions <- list(c(0.35, 0.975), c(0.21, 0.975))
for(i in seq_along(vrs)) { # iterate from 1 to the number of variables
  # Get data for this iteration
  var_selected <- vrs[i]
  color_selected <- col_line[i]
  # Call the plotting function
  p <- plot_gam_effect(
    model_dtw, var_selected, color_selected,
    y_label = if (i == 1) "Predicted Value\n(DBA Centroid 1)" else NULL
  )
  # Add the tag (LETTERS)
  plot_list[[var_selected]] <- p + labs(tag = LETTERS[i]) +
    theme(
      plot.tag.position = tag_positions[[i]],
      plot.tag = element_text(
        face = "bold", size = 20)
      )
}
# Saving figure elements
plot_GAM_c1 <- plot_list
save(plot_GAM_c1,
  file = paste0("outputs/files/",'Fig_8_AB_c1_GAM.RData')
  )
# --- Centroid 4
cat("\n\n =============== Centroid 4 ===============")
model_dtw <- gam(
  (c4 ~ s(sam, k = 4, bs = "cr") + s(kd75, k = 4, bs = "cr")),
  family = quasi(link="identity", variance="constant"),
  data = env_fish_lag_dtw[[11]], na.action = na.omit, method = "REML"
  )
summary(model_dtw)
de_dtw <- round((1 - model_dtw[["deviance"]] / model_dtw[["null.deviance"]]) * 100, digits = 2)
concurvity(model_dtw)
shapiro.test(model_dtw$residuals)
var_contrib_from_full_gam(model_dtw)
# Figure 
vrs <- names(model_dtw[["var.summary"]]) # 
col_line <- c("#FFB6C1", "#E69F00") # sam, kd
plot_list <- list() # We create the list to store the plots
tag_positions <- list(c(0.35, 0.975), c(0.21, 0.975))
for(i in seq_along(vrs)) { # iterate from 1 to the number of variables
  # Get data for this iteration
  var_selected <- vrs[i]
  color_selected <- col_line[i]
  # Call the plotting function
  p <- plot_gam_effect(
    model_dtw, var_selected, color_selected,
    y_label = if (i == 1) "Predicted Value\n(DBA Centroid 4)" else NULL
  )
  # Add the tag (LETTERS)
  plot_list[[var_selected]] <- p + labs(tag = LETTERS[i+7]) +
    theme(
      plot.tag.position = tag_positions[[i]],
      plot.tag = element_text(
        face = "bold", size = 20)
      )
}
# Saving figure elements
plot_GAM_c4 <- plot_list
save(plot_GAM_c4,
  file = paste0("outputs/files/",'Fig_8_HI_c4_GAM.RData')
  )

# Checking model residuals for DFA with highest DE, and plotting
# --- State 1
cat("\n\n =============== Common hidden trend 1 ===============")
model_dfa <- gam(
  (s1 ~ s(soi, k = 4, bs = "cr") + s(kd50, k = 4, bs = "cr")),
  family = quasi(link="identity", variance="constant"),
  data = env_fish_lag_dfa[[10]], na.action = na.omit, method = "REML"
  )
summary(model_dfa)
de_dfa <- round((1 - model_dfa[["deviance"]] / model_dfa[["null.deviance"]]) * 100, digits = 2)
concurvity(model_dfa)
shapiro.test(model_dfa$residuals)
var_contrib_from_full_gam(model_dfa)
# Figure
vrs <- names(model_dfa[["var.summary"]]) # 
col_line <- c("#492050", "#E69F00") # soi, kd
plot_list <- list() # We create the list to store the plots
tag_positions <- list(c(0.35, 0.975), c(0.21, 0.975))
for(i in seq_along(vrs)) { # iterate from 1 to the number of variables
  # Get data for this iteration
  var_selected <- vrs[i]
  color_selected <- col_line[i]
  # Call the plotting function
  p <- plot_gam_effect(
    model_dfa, var_selected, color_selected,
    y_label = if (i == 1) "Predicted Value\n(DFA Common trend 1)" else NULL
  )
  # Add the tag (LETTERS)
  plot_list[[var_selected]] <- p + labs(tag = LETTERS[i]) +
    theme(
      plot.tag.position = tag_positions[[i]],
      plot.tag = element_text(
        face = "bold", size = 20)
      )
}
# Saving figure elements
plot_GAM_s1 <- plot_list
save(plot_GAM_s1,
  file = paste0("outputs/files/",'Fig_9_AB_s1_GAM.RData')
  )
# --- State 2
cat("\n\n =============== Common hidden trend 2 ===============")
model_dfa <- gam(
  (s2 ~ s(sss25, k = 4, bs = "cr") + s(sst25, k = 4, bs = "cr")),
  family = quasi(link="identity", variance="constant"),
  data = env_fish_lag_dfa[[10]], na.action = na.omit, method = "REML"
  )
summary(model_dfa)
de_dfa <- round((1 - model_dfa[["deviance"]] / model_dfa[["null.deviance"]]) * 100, digits = 2)
concurvity(model_dfa)
shapiro.test(model_dfa$residuals)
var_contrib_from_full_gam(model_dfa)
# --- State 3
cat("\n\n =============== Common hidden trend 3 ===============")
model_dfa <- gam(
  (s3 ~ s(kd75, k = 4, bs = "cr") + s(sss75, k = 4, bs = "cr")),
  family = quasi(link="identity", variance="constant"),
  data = env_fish_lag_dfa[[7]], na.action = na.omit, method = "REML"
  )
summary(model_dfa)
de_dfa <- round((1 - model_dfa[["deviance"]] / model_dfa[["null.deviance"]]) * 100, digits = 2)
concurvity(model_dfa)
shapiro.test(model_dfa$residuals)
var_contrib_from_full_gam(model_dfa)

```

# **Response plots of explanatory variables retained in the best-fitting GLM and GAM models for both centroids (Figure 8) and common hidden trends (Figure 9)**

```{r, fig.width=12, fig.height=9}
#  CENTROIDS ----
#path_plots <- paste0("out_dir_fig")
load(paste0(out_dir_files, '/Fig_8_AB_c1_GAM.RData'))
load(paste0(out_dir_files, '/Fig_8_CD_c2_GLM.RData'))
load(paste0(out_dir_files, '/Fig_8_EFG_c3_GLM.RData'))
load(paste0(out_dir_files, '/Fig_8_HI_c4_GAM.RData'))
nc = 4 # number of columns
row1 <- arrangeGrob(grobs = plot_GAM_c1, nrow = 1, ncol = nc)
row2 <- arrangeGrob(grobs = plot_GLM_c2, nrow = 1, ncol = nc)
row3 <- arrangeGrob(grobs = plot_GLM_c3[c(1,2,4)], nrow = 1, ncol = nc)
row4 <- arrangeGrob(grobs = plot_GAM_c4, nrow = 1, ncol = nc)
# Save figure
file_path <- file.path(out_dir_fig, "Fig_8")
# --- as PDF
pdf(paste0(file_path, ".pdf"), width = 12, height = 9, colormodel = "cmyk")
par(mfcol=c(1,1))
final_dtw <- grid.arrange(grobs = list(row1, row2, row3, row4), nrow = 4,
                          width = c(1, 1, 1))
grid::grid.draw(final_dtw)
invisible(dev.off())
# --- as PNG
ragg::agg_png(paste0(file_path, ".png"), width = 12, height = 9, units = "in", res = 600)
grid::grid.draw(final_dtw)
invisible(dev.off())
# HTML output
file_png <- paste0(file_path,".png")
if (knitr::is_html_output() && file.exists(file_png)) {
  knitr::include_graphics(file_png)
}
```

**Figure 8.-** Response plots for the best-fitting GLMs and GAMs obtained for the DTW-derived centroids 1–4. Each row corresponds to a centroid, and each panel shows the partial effect of the anomaly of one explanatory variable (or interaction) on its predicted values, based on the selected model for that centroid (see details about lags and percentiles in Tables 2 and 3). Solid lines represent the estimated mean effect, and shaded areas denote the 95% confidence intervals. Rug marks along the x-axis indicate the distribution of observed covariate values. Abbreviations of environmental explanatory variables are defined in Table 1

```{r}
#  TRENDS ----
load(paste0(out_dir_files, '/Fig_9_AB_s1_GAM.RData'))
load(paste0(out_dir_files, '/Fig_9_CDE_s2_GLM.RData'))
load(paste0(out_dir_files, '/Fig_9_FGHI_s3_GLM.RData'))
row1 <- arrangeGrob(grobs = plot_GAM_s1, nrow = 1, ncol = nc)
row2 <- arrangeGrob(grobs = plot_GLM_s2, nrow = 1, ncol = nc)
row3 <- arrangeGrob(grobs = plot_GLM_s3, nrow = 1, ncol = nc)
# Save figure
file_path <- file.path(out_dir_fig, "Fig_9")
# --- as PDF
pdf(paste0(file_path, ".pdf"), width = 12, height = 9, colormodel = "cmyk")
par(mfcol=c(1,1))
final_dfa <- grid.arrange(grobs = list(row1, row2, row3), nrow = 4,
                          )
grid::grid.draw(final_dfa)
invisible(dev.off())
# --- as PNG
ragg::agg_png(paste0(file_path, ".png"), width = 12, height = 9, units = "in", res = 600)
grid::grid.draw(grid.arrange(grobs = list(row1, row2, row3), nrow = 3,
                          ))
invisible(dev.off())
# HTML output
file_png <- paste0(file_path,".png")
if (knitr::is_html_output() && file.exists(file_png)) {
  knitr::include_graphics(file_png)
}
```

**Figure 9.-** Response plots for the best-fitting GLMs and GAMs obtained for the DFA-derived common hidden trends 1–3. Each row corresponds to a common trend, and each panel shows the partial effect of the anomaly of one explanatory variable on its predicted values, based on the selected model for that common trend (see details about lags and percentiles in Tables 2 and 3). Solid lines represent the estimated mean effect, and shaded areas denote the 95% confidence intervals. Rug marks along the x-axis indicate the distribution of observed covariate values. Abbreviations of environmental explanatory variables are defined in Table 1.

# **Previous section**

::: callout-note
## To return to previous analyses, focusing on DTW and DFA techniques:

[Go to Code 1 of 2: DTW and DFA analyses](1_dtw_dfa_analyses.html){.btn .btn-outline-primary role="button"}
:::
